---
title: "BetaDiversity"
output: html_document
---


Load in taxonomy phyloseq object, run beta diversity, plot  results, and run statistics.

```{r Load data and libraries}
library(tidyverse)
library(phyloseq)
library(vegan)

# Load object
ps = readRDS('../datasets_and_results/phyloseq_taxonomy.rds')

# rarefy the data:
psrare = ps %>% rarefy_even_depth(sample.size = 500, rngseed = 420)
```

```{r calculate bray-curtis distances}
set.seed(420)
# Calculate the distance between each sample pair
ps_bray = phyloseq::distance(psrare, method = "wunifrac")
View(as.matrix(ps_bray)) # Every pairwise comparison included
```

Apply multidimensional scaling (MDS). This takes the distance matrix and reduces it to a few dimensions. 

```{r normalize and plot}
set.seed(420)
# MDS scaling
mds = metaMDS(ps_bray)

# Extract data
mds_data = mds$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F)
head(mds_data)

# Plot it
p = mds_data %>%
  ggplot(aes(MDS1,MDS2,color = p_order)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size=18)
p
```

Run PERMANOVAs using the adonis() function from the vegan package.

```{r PERMANOVA on p_order}
set.seed(420)
metadata = sample_data(psrare) %>% data.frame() # to make the next lines easier

# Single variable
stats_univar = adonis2(ps_bray ~ p_order, data = metadata)

# Two variables - adding Diet as a covariable
# by='margin' is important here - it gives the stats for each variable separately
stats_multivar = adonis2(ps_bray ~ p_order + Diet, data = metadata, by = 'margin') 

# Combine into a table
stats = bind_rows('Univariate' = stats_univar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  'Multivariate' = stats_multivar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') # First column is Model, either Uni or Multi

# Add some formattings
stats = stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval)) # Remove the irrelevant lines

stats 
```

Subset the distance matrix object, then compare specific groups that are shown in the beta diversity plot

```{r attempt 2 - subset distance matrix}
set.seed(420)
# Select two orders to compare
to_compare = c('Perciformes','Labriformes')
# Labriformes Centrarchiformes Blenniiformes

# List samples that belong to those orders
samples_to_keep = metadata %>% filter(p_order %in% to_compare) %>% rownames()

# Subset the distance matrix to only include those samples
ps_bray_sub = as.matrix(ps_bray)[samples_to_keep,samples_to_keep] %>% as.dist()

# Run a PERMANOVA on just those two oders
stats_univar = adonis2(ps_bray_sub ~ p_order,
                       data = metadata %>% filter(p_order %in% to_compare))

stats_univar 
```

P values are usually added manually to beta diversity plots, or are added to the figure legend.

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

We've established that body site is a major driver of beta diversity. In particular, most of the differences seem to be driven by the fact that the gut samples ar dramatically different from everything else.

What if you don't care about the gut samples at all? Beta diversity calculations are influenced by every sample included in the dataset, so even if you remove the gut samples after calculating distances, they're still going to have a huge influence on the results.

In this case, it's best to simply remove the gut samples from the phyloseq object, then recalculate distances and MDS.

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

Let's save our beta diversity plot, as well as our pre-subsetting statistics.

```{r save outputs}
# Save plot
ggsave('../Results - Beta Diversity.jpeg',
       plot = p,
       height= 4, width = 6)

# Save statistics as an Excel file
writexl::write_xlsx(list('Beta Diversity' = stats),
                    '../Results - Beta Diversity.xlsx')
```